Evolution of ORM Systems: Formal Foundations

### 2013-01-30 doc/description.tex
case study summary, merge transformation fixed

\paragraph{Merge of Classes} There occur situations when there is a need to merge two classes and the inlining transformation cannot be used because of name collisions among properties in such a case the $merge$ transformation can be used. It can be used in case structure of both classes (respectively tables) is equals. The transformation creates new rows in the new table - the primary keys from origin instances are replaced by a new ones.
\begin{align*}
&	\mathbf{Merge} : Label \times Label \times Label \times State \rightarrow State \\
&	\Phi(s, Merge(l_1, l_2, l_n) = \inference{
		\exists c_1, c_2 \in classes(application(s)) \wedge 
		equals(c_1, c_2)}{\begin{gathered}
			application(s) \\ \circ addClass(cl(ln, properties(c_1), associations(c_1)) \\
			\circ (\forall c \in classes (application(s)), a \in associations(c) \\
			reference(a) = c_1 \vee reference(a) = c_2 : \\
			reference(a) = l_n) \\
			\circ renClass(l_1) \circ renClass(l_2) \end{gathered}}
 \\\\
&	\Psi(s, merge(l_1, l_2, l_n)) = \inference{\exists t_1, t_2 \in tables(database(s)) : \\ name(t_1) = l_1 \wedge name(t_2) = l_2 \\ \wedge columns(t_1) = columns(t_2)}{\begin{gathered}
	 database(s) \\ \bullet addTable(l_n, pk("Id", s_1), columns(t_1), foreignKeys(t_1)) \\ 
\bullet (\forall r \in instances(t_1, database(s)) \cup instances(t_2, database(s)) :  \\ n =  row(l_n, next(seq(key(l_n))), pairs(r)), insertRow(l_n, n) \\ 
\bullet (\forall k \in data(database(s)), f \in foreignKeys(k) \\ \wedge reference(f) = reference(r) \\ \wedge value(findPair(name(f), k)) = pkValue(r) : \\value(findPair(name(f), k)) = pkValue(n))) \\  \bullet dropTableAndData(l_1) \bullet dropTableAndData(l_2)
 \end{gathered}} \\
 & l_1, l_2, l_n \in Label, s \in State
\end{align*}

@@

\begin{figure}
\begin{subfigure}[b]{0.5\textwidth}
	\includegraphics[width=\textwidth]{./images/case_app_1}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
	\includegraphics[width=\textwidth]{./images/case_db_1}
\end{subfigure}
\end{figure}

The first  step of is to extract two temporary classes representing address of a $Person$ or of a $LegalPerson$. Temporary classes has to be connected with the origin classes by an association.
\begin{align*}
& state(S) \oplus Extract("Person", "Address\_tmp1", "street", "city", "zip") \\
& \oplus AddAssociation("Person", assoc("address", "Address\_tmp1", 1, 1),\\ 
& \{(x,y)| x \in instances("Person", database(S)), \\
& y \in instances("Address\_tmp1", database(S)) : pkVal(x) = pkVal(y) \}) \\
& \oplus Extract("LegalPerson", "Address\_tmp2", "street", "city", "zip")  \\
& \oplus AddAssociation("LegalPerson", assoc("residentialAddress", "Address\_tmp2", 1, 1),\\ 
& \{(x,y)| x \in instances("LegalPerson", database(S)), \\
& y \in instances("Address\_tmp2", database(S)) : pkVal(x) = pkVal(y) \}) \\
\end{align*}

\begin{figure}
\begin{subfigure}[b]{0.5\textwidth}
	\includegraphics[width=\textwidth]{./images/case_app_5}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
	\includegraphics[width=\textwidth]{./images/case_db_5}
\end{subfigure}
\end{figure}
%Next two evolution steps are bit intuitive but necessary for the next evolution.  then the temporary class can be merged together and finally the temporary class has to be renamed to $Address$:
The two temporary classes can be merged together as they have the same structure. All associations pointing to one of the temporary classes are redirected to the newly created class. Finally the associations are renamed.
\begin{align*}
%
%& \oplus AddAssociation("Person", \\
%& assoc("address", "Address\_tmp2", 1, 1), \varnothing) \\
%& \oplus AddAssociation("LegalPerson", \\
%& assoc("residentialAddress", "Address\_tmp1", 1, 1),\varnothing) \\
%
& \oplus Merge("Address\_tmp1", "Address\_tmp2", "Address") \\
\end{align*}
Finally the associations are renamed and the evolution end.
\begin{align*}
& \oplus RenameAssociation("Person", "Address\_tmp1", "Address")\\
& \oplus RenameAssociation("LegalPerson", "Address\_tmp2". "Address")\\
\end{align*}

\begin{figure}
\begin{subfigure}[b]{0.5\textwidth}
	\includegraphics[width=\textwidth]{./images/case_app_9}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
	\includegraphics[width=\textwidth]{./images/case_db_8}
\end{subfigure}
\end{figure}

>>>>

\paragraph{Merge of Classes} There occur situations when there is a need to merge two classes and the inlining transformation cannot be used because of name collisions among properties in such a case the $merge$ transformation can be used. It can be used in case structure of both classes (respectively tables) is equals. The transformation creates new rows in the new table - the primary keys from origin instances are replaced by a new ones.
\begin{align*}
&	\mathbf{Merge} : Label \times Label \times Label \times State \rightarrow State \\
&	\Phi(s, Merge(l_1, l_2, l_n) = \inference{
		\exists c_1, c_2 \in classes(application(s)) \wedge 
		equals(c_1, c_2)}{\begin{gathered}
			application(s) \\ \circ addClass(cl(ln, properties(c_1), associations(c_1)) \\
			\circ (\forall c \in classes (application(s)), a \in associations(c) \\
			reference(a) = c_1 \vee reference(a) = c_2 : \\
			reference(a) = l_n) \\
			\circ renClass(l_1) \circ renClass(l_2) \end{gathered}}
 \\\\
&	\Psi(s, merge(l_1, l_2, l_n)) = \inference{\exists t_1, t_2 \in tables(database(s)) : \\ name(t_1) = l_1 \wedge name(t_2) = l_2 \\ \wedge columns(t_1) = columns(t_2)\\
\forall f \in (foreignKeys(t_1) \cup foreignKey(t_2)) \\ \setminus (foreignKey(t_1) \cap foreignKey(t_2)) \\ : NOTNULL \not \in constraints(f)  }{\begin{gathered}
	 database(s) \\ \bullet addTable(l_n, pk("Id", s_1), columns(t_1), foreignKeys(t_1)) \\ 
\bullet (\forall r \in instances(t_1, database(s)) \cup instances(t_2, database(s)) :  \\ n =  row(l_n, next(seq(key(l_n))), pairs(r)), insertRow(l_n, n) \\ 
\bullet (\forall k \in data(database(s)), f \in foreignKeys(k) \\ \wedge reference(f) = reference(r) \\ \wedge value(findPair(name(f), k)) = pkValue(r) : \\value(findPair(name(f), k)) = pkValue(n))) \\  \bullet dropTableAndData(l_1) \bullet dropTableAndData(l_2)
 \end{gathered}} \\
 & l_1, l_2, l_n \in Label, s \in State
\end{align*}

@@

\begin{figure}
\begin{subfigure}[b]{0.5\textwidth}
	\includegraphics[width=\textwidth]{./images/case_app_1}
	\caption{The application model.}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
	\includegraphics[width=\textwidth]{./images/case_db_1}
	\caption{The database model.}
\end{subfigure}
	\caption{The initial state of the example used in the case study. There is a repetition of information structure in the $Person$ and $LegalPerson$ class.}
	\label{fig:case1}
\end{figure}

The first  step of is to extract two temporary classes representing address of a $Person$ or of a $LegalPerson$. Temporary classes has to be connected with the origin classes by an association. The result is shown in Figure \ref{fig:case2}.
\begin{align*}
S_1 &= state(S) \oplus Extract("Person", "Address\_tmp1", "street", "city", "zip") \\
& \oplus AddAssociation("Person", assoc("address", "Address\_tmp1", 1, 1),\\ 
& \{(x,y)| x \in instances("Person", database(S)), \\
& y \in instances("Address\_tmp1", database(S)) : pkVal(x) = pkVal(y) \}) \\
& \oplus Extract("LegalPerson", "Address\_tmp2", "street", "city", "zip")  \\
& \oplus AddAssociation("LegalPerson", assoc("residentialAddress", "Address\_tmp2", 1, 1),\\ 
& \{(x,y)| x \in instances("LegalPerson", database(S)), \\
& y \in instances("Address\_tmp2", database(S)) : pkVal(x) = pkVal(y) \}) \\
\end{align*}

\begin{figure}
\begin{subfigure}[b]{0.5\textwidth}
	\includegraphics[width=\textwidth]{./images/case_app_5}
	\caption{The application model.}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
	\includegraphics[width=\textwidth]{./images/case_db_5}
	\caption{The database model.}
\end{subfigure}
	\caption{The state of the software after the information are extracted into separated classes and associations are created.}
	\label{fig:case2}
\end{figure}
%Next two evolution steps are bit intuitive but necessary for the next evolution.  then the temporary class can be merged together and finally the temporary class has to be renamed to $Address$:
The two temporary classes can be merged together as they have the same structure. All associations pointing to one of the temporary classes are redirected to the newly created class.
\begin{align*}
%
%& \oplus AddAssociation("Person", \\
%& assoc("address", "Address\_tmp2", 1, 1), \varnothing) \\
%& \oplus AddAssociation("LegalPerson", \\
%& assoc("residentialAddress", "Address\_tmp1", 1, 1),\varnothing) \\
%
S_2 &= S_1 \oplus Merge("Address\_tmp1", "Address\_tmp2", "Address") \\
\end{align*}
Finally the associations are renamed and the evolution end.
\begin{align*}
S_3 &= S_2  \oplus RenameAssociation("Person", "Address\_tmp1", "Address")\\
& \oplus RenameAssociation("LegalPerson", "Address\_tmp2". "Address")\\
\end{align*}

\begin{figure}
\begin{subfigure}[b]{0.5\textwidth}
	\includegraphics[width=\textwidth]{./images/case_app_9}
	\caption{The application model.}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
	\includegraphics[width=\textwidth]{./images/case_db_8}
	\caption{The database model.}
\end{subfigure}
	\caption{The final state of the case study example. There is one class containing information about address for the whole software and data were migrated.}
\end{figure}

The scenario was completed successfully, because there were no constraints defined in the database and the data was suitable for the evolution there were no complication. The example of evolution presented in case study shows the possibilities of defined transformation. The example itself is not large, nevertheless it shows advanced refactoring cases, which are very often used in real software development. We believe that composition of proposed transformation can be used even in case of large data evolution cases. It is obvious that usage of proposed set of transformations only will cause very large scripts describing data evolution, therefore a mechanism for creating templates (reusable evolution scripts).
